---
title: "R Markdown Example File"
author: "Andrew Zieffler"
date: "September 21, 2014"
output: 
  ioslides_presentation:
    toc: true
bibliography: myBibliography.bib
csl: apa-single-spaced.csl
---

# Tables

First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell


# Example 1

```{r}
summary(cars)
```

# Example 2

```{r myChunk, echo=FALSE}
summary(cars)
```

# Example 3

```{r fig1, out.width='600px', out.height='600px',fig.cap='Here is my figure caption.', fig.align='center'}
library(ggplot2)
ggplot(data=cars, aes(x=speed, y=dist)) +
  geom_point() +
  theme_bw()  
```

# Inline Code Chunks

Here is an example of an inline code chunk. The mean speed is `r mean(cars$speed)`.  

# Display Equation

$$
\hat{Y_{ij}} = \beta_{0} + \beta_{1}(X_{1}) + \epsilon_{ij}
$$  

# Citations

Here is some text and a citation [@agresti; @algina].

Here is some text and a citation [see @agresti; @algina, pp. 317].

Agresti [-@agresti] says that this is cool stuff .

# New Slide

- Here is bullet 1
- Here is bullet 2

# References

